Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-04-22 paper/paper.tex
morer stuffer

RL is a problem-solving technique in which, through interactions with an unknown \textit{environment},
an agent learns to reach goals, optimize reward signals, and minimize costs
in an iterative trial-by-error fashion. See Fig. \ref{fig:rl}.
At a given moment, each taken \textit{action} $a$ triggers an \textit{partially observed} change
in the \textit{state} of the environment
$s \rightarrow s'$, accompanied by environmental feedback signals as \textit{reward}
$r = r(s, a)$ collected by the agent (Fig. \ref{fig:rl}).
If the collected reward outcome yields a negative value it can be
more naturally interpreted as \textit{punishment}.
In this view, the environment is partially controlled by
the action of the agent and the reward can be thought
of as satisfaction -- or aversion -- accompanying the execution of
an action.

Action may be delayed to achieve substantial increases in expected reward
that is expected to grow with time.
The environment is generally taken as \textit{stochastic},
that is, changing in random ways. In addition,
the environment is only \textit{partially observed}
in the sense that only part of the current state is observed by
the agent \citep{starkweather2017dopamine}.
Assuming volatility of the environment
is realistic in a computational model which sets out
to explain DMN function.
% XXX: the following passage is already in the introduction
% Indeed, it is known that the more the external world is predictable,
% the more mental activity becomes detached from the actual sensory inputs
% \citep{antrobus1966studies, pope1978regulation},
% the more stimulus-independent thoughts occur, and
% the higher the neural activity in the DMN.
% Conversely, the more the currently executed task is unknown and unpracticed,
% the lower the DMN activity
% \citep{filler1973daydreaming, teasdale1995stimulus}.
% %
% These ``offline'' processes may however contribute to
% optimizing control of the organism.
% Such a framework of DMN funtion based on reinforcement learning
% can naturally embed human behavior
% in the tension between exploitative action with immediate gains and
% explorative action with longer-term reward schedules
% \citep{dayan2008decision}.
We argue that a functional account of the DMN based on RL
can naturally embed human behavior
in the tension between exploitative action with immediate gains and
explorative action with longer-term reward outcomes
\citep{dayan2008decision}.

@@

\paragraph{Model-free reinforcement learning.}
Model-free RL has had great success in real-world problems including robotics\citep{abbeel2004,andrewng}, computer game play, \citep{mnih2015,silver2016mastering},
etc., just to name a few. Such a learning scheme can be
plausibly realized in the human brain. Indeed,
a core property of human intelligence has been proposed underly to improve
on an expected utility outcomes as a strategy for action choice in uncertain
environments \citep{gershman2015computational}, a situation perfectly captured by the formalism of POMDPs.
It has also long been proposed \citep{dayan2008decision} that there
is a rather direct mapping of model-free RL learning algorithms
onto aspects of the brain.
The neurotransmitter dopamine could serve
as a ``teaching signal'' to better estimate value associations
and action policies by controlling
synaptic plasticity at targets such as the ventral striatum.
In contrast, \textit{model-based RL} would start off with some mechanistic assumptions about the dynamics of the world.
These can be assumptions about the physical laws governing the agent's world or constraints on the state space and transitions between states.
For instance, if a rat in a maze knows
that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
%

@@

\begin{equation}
  \pi^{\text{greedy}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
by querying an approximate version $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$. For instance, a simplistic linear model with a kernel $\phi$ would be of the
form $\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta$, where
$\phi(s,a)$ would represent a high-level representation of the state-action pairs
$(s,a)$, as proposed by \citep{songNIPS2016} for example.
%  and Parr et al. 2003 (JMLR)].
In the DMN, The dmPFC (BA9) would implement such a greedy lookup over the action space.
The tensor $\theta$ encodes the model
  parameters and would correspond to synaptic weights and / or connection strengths between
  DMN regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the greedy policy \eqref{eq:policy}.
  The oracle Q-values $Q(s, a|\theta)$ tell the DMN the (expected) usefulness of taking an action a in state / situation  $s$.
  The DMN, and in particular its dmPFC node, would then contribute to the choice, at a given $s$, of an action $a$ which maximizes this tabulated values. This mapping from states to actions defines a matrix, which can be conveniently called a "policy matrix". Learning consists in starting from a given instantiation of such a table and tweaking it around while using it to take actions which take the agent to different areas of the table, and so forth.

@@

% \paragraph{Parameter update.}
\begin{equation}
  \delta \theta_{k} \propto -\nabla_{\theta_{k}}\mathcal L(\theta_{k})
  = -\mathbb E_{(s, a, r, s') \sim \mathcal D}[\underbrace{(\tilde{Q}(s, a|\theta_{k}) - y_k)}_{\text{prediction error}}
    \underbrace{\nabla_{\theta_{k}}\tilde{Q}(s, a|\theta_{k})}_{\text{aversion}}].
  \label{eq:oracle}
\end{equation}
Over time (i.e., in the limit $k \rightarrow \infty$),
the thus generated approximating sequence of
$Q$-value functions provably\footnote{The proof is standard and relies on the Banach Fixed-Point Theorem applied to the Bellman transform.} $Q$-value functions $\tilde{Q}(.,.|\theta_0) \rightarrow \tilde{Q}(.,.|\theta_1) \rightarrow \tilde{Q}(.,.|\theta_2) \rightarrow \ldots$
are attracted and absorbed by the optimal $Q$-value function $Q^*$ defined implicitly by the Bellman equation \eqref{eq:bellman}.

>>>>

RL is a problem-solving technique in which, through interactions with an unknown \textit{environment},
an agent learns to reach goals, optimize reward signals, and minimize costs
in an iterative trial-by-error fashion. See Fig. \ref{fig:rl}.
At a given moment, each taken \textit{action} $a$ triggers an \textit{partially observed} change
in the \textit{state} of the environment
$s \rightarrow s'$, accompanied by environmental feedback signals as \textit{reward}
$r = r(s, a)$ collected by the agent (Fig. \ref{fig:rl}).
If the collected reward outcome yields a negative value it can be
more naturally interpreted as \textit{punishment}.
In this view, the environment is partially controlled by
the action of the agent and the reward can be thought
of as satisfaction -- or aversion -- accompanying the execution of
an action.
Action may be delayed to achieve substantial increases in expected reward
that is expected to grow with time.

The environment is generally taken as \textit{stochastic},
that is, changing in random ways. In addition,
the environment is only \textit{partially observed}
in the sense that only part of the current state is observed by
the agent \citep{starkweather2017dopamine}.
Assuming volatility of the environment
is realistic in a computational model which sets out
to explain DMN function.
% XXX: the following passage is already in the introduction
% Indeed, it is known that the more the external world is predictable,
% the more mental activity becomes detached from the actual sensory inputs
% \citep{antrobus1966studies, pope1978regulation},
% the more stimulus-independent thoughts occur, and
% the higher the neural activity in the DMN.
% Conversely, the more the currently executed task is unknown and unpracticed,
% the lower the DMN activity
% \citep{filler1973daydreaming, teasdale1995stimulus}.
% %
% These ``offline'' processes may however contribute to
% optimizing control of the organism.
% Such a framework of DMN funtion based on reinforcement learning
% can naturally embed human behavior
% in the tension between exploitative action with immediate gains and
% explorative action with longer-term reward schedules
% \citep{dayan2008decision}.
We argue that a functional account of the DMN based on RL
can naturally embed human behavior
in the tension between exploitative action with immediate gains and
explorative action with longer-term reward outcomes
\citep{dayan2008decision}.

@@

\paragraph{Model-free reinforcement learning.}
Model-free RL has had great success in real-world problems including robotics\citep{abbeel2004,andrewng}, computer game play, \citep{mnih2015,silver2016mastering},
etc., just to name a few. Such a learning scheme can be
plausibly realized in the human brain. Indeed,
a core property of human intelligence has been proposed to underly improvement
of expected utility outcomes as a strategy for action choice in uncertain
environments \citep{gershman2015computational}, a situation perfectly captured by the formalism of POMDPs.
It has also long been proposed \citep{dayan2008decision} that there
is a rather direct mapping of model-free RL learning algorithms
onto aspects of the brain.
The neurotransmitter dopamine could serve
as a ``teaching signal'' to better estimate value associations
and action policies by controlling
synaptic plasticity at targets such as the ventral striatum.
In contrast, \textit{model-based RL} would start off with some mechanistic assumptions about the dynamics of the world.
These can be assumptions about the physical laws governing the agent's world or constraints on the state space and transitions between states.
For instance, if a rat in a maze knows
that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
%

@@

\begin{equation}
  \pi^{\text{greedy}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
by querying an approximate version $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$. For instance, a simplistic linear model with a kernel $\phi$ would be of the
form $\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta$, where
$\phi(s,a)$ would represent a high-level representation of the state-action pairs
$(s,a)$, as proposed by \citep{songNIPS2016} for example.
%  and Parr et al. 2003 (JMLR)].
In the DMN, The dmPFC (BA9) would implement such a greedy lookup over the action space.
The tensor $\theta$ encodes the model
  parameters and would correspond to synaptic weights and connection strengths between
  DMN regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the greedy policy \eqref{eq:policy}.
  The oracle Q-values $Q(s, a|\theta)$ tell the DMN the (expected) usefulness of taking an action a in state / situation  $s$.
  The DMN, and in particular its dmPFC node, would then contribute to the choice, at a given $s$, of an action $a$ which maximizes this tabulated values. This mapping from states to actions defines a matrix, which can be conveniently called a "policy matrix". Learning consists in starting from a given instantiation of such a table and tweaking it around while using it to take actions which take the agent to different areas of the table, and so forth.

@@

% \paragraph{Parameter update.}
\begin{equation}
  \delta \theta_{k} \propto -\nabla_{\theta_{k}}\mathcal L(\theta_{k})
  = -\mathbb E_{(s, a, r, s') \sim \mathcal D}[\underbrace{(\tilde{Q}(s, a|\theta_{k}) - y_k)}_{\text{prediction error}}
    \underbrace{\nabla_{\theta_{k}}\tilde{Q}(s, a|\theta_{k})}_{\text{aversion}}].
  \label{eq:oracle}
\end{equation}
Over time (i.e., in the limit $k \rightarrow \infty$),
the thus generated approximating sequence of
$Q$-value functions provably\footnote{The proof is standard and relies on the Banach Fixed-Point Theorem applied to the Bellman transform.}
$\tilde{Q}(.,.|\theta_0) \rightarrow \tilde{Q}(.,.|\theta_1) \rightarrow \tilde{Q}(.,.|\theta_2) \rightarrow \ldots$
are attracted and absorbed by the optimal $Q$-value function $Q^*$ defined implicitly by the Bellman equation \eqref{eq:bellman}.
